Leadership across cultures and contexts: an exploratory analysis

### 2020-06-22 leadership_across_cultures_contexts.Rmd
general text reduction

Anthropologists and sociologists realized early on that leadership and followership were critical to understanding human psychology, social organization, and culture [e.g., @mumford_origins_1906; @firth1927economic; @morgan_ancient_1877; @myres_influence_1917]. Leaders are documented among every ethnographically observed society [@brown_human_1991; @lewis_leaders_1974], and in diverse contexts, such as kin groups [@dussart2000politics], ritual [@singh_cultural_2017], economic groups [@macfarlan_competitive_2012], conflicts between groups [@glowacki_evolutionary_2017], and in nonindustrial political groups [@cohen_comparative_1967]. For review, see @garfield_review_2019.

Despite decades of research and theorizing on leadership in numerous disciplines, to our knowledge there is no systematic investigation of leader functions and qualities, and the costs and benefits of leadership and followership, across a representative sample of nonindustrial societies. Moreover, outside of anthropology, most research and theorizing on leadership is based on data from Western or Westernized, postindustrial societies [@von_rueden_leadership_2015].

<!-- WEIRD PEOPLE PROBLEMS -->
There is now a broad consensus on the importance of data from the full range of human cultural diversity. Since the seminal WEIRD people paper [@henrich_weirdest_2010], many social scientists have taken up the challenge to obtain data from "non-WEIRD" populations. Dichotomizing populations and their psychologies and social dynamics as WEIRD or non-WEIRD, however, is a mistake.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] As anthropology has conclusively demonstrated, there is enormous diversity across societies that does not remotely resemble a dichotomy.

@@

## The functions of leadership across cultures and contexts
<!-- INTRO -->
@roscoe_social_2009 proposed leadership in small-scale societies is structured by the joint demands of facilitating within-group conflict resolution and cooperation, and signaling coalitional strength to rival groups. Among many small-scale societies, which generally lack institutionalized political structures, leadership at the community level is typically informal and leaders emerge in response to group demands [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. Theoretical literature suggests these demands most commonly arise in the contexts of collective actions, resolving conflicts between families, and punishing norm violations [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2020]. Leaders also play important roles in ritual contexts, such as in initiations [@singh_cultural_2017], facilitating marriages [@walker_evolutionary_2011], and organizing feasts [@wiessner_food_1996]. Between-group interactions, including both conflict and cooperation, are also common among hunter-gatherers and other nonindustrial societies [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012].

@@

<!-- MANAGEMENT FUNCTION IN MORE CULTUARLL COMPLEX -->
Stronger authority of community leaders, leadership beyond the community level, and institutionalized managerial roles are associated with increased sedentarization, defensibility of resources, and social stratification [@johnson_evolution_1987; @johnson_organizational_1982; @kaplan_evolutionary_2009; @ames_hierarchies_1985; @panchanathan_indirect_2004]. Among horticulturalists and agriculturalists, managerial leadership can promote more efficient use of shared resources such as water reserves and cooperative labor [e.g., @von_rueden_leadership_2014]. Pastoralists and coastal populations often face similar pressures concerning grazing lands and fishing access [e.g., @widmer_evolution_1988; @stevens_sherpa_1990]. 
<!-- In the wake of economic intensification, the functions of leadership becomes increasingly associated with managing economic systems and distribution of resources.  -->
Defensibility of critical resources creates increased opportunities for resource management and is often associated with increased territoriality and inter-group conflicts [@glowacki_evolutionary_2017]. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military operations [@lopez_evolutionary_2017].

@@

<!-- STATUS VS LEADERSHIP -->
Leaders have disproportionate group influence, whereas status is often conceptualized as a relative indicator of the social value of the individual by the group and of their relative access to contested resources [e.g., @blader_whats_2014; @von_rueden_roots_2014]. Leaders are expected to be high status [e.g., @von_rueden_leadership_2015; @van_vugt_genes_2020; @cheng_dominance_2019; @cheng_psychology_2014; @buss_human_2020]. The causal relationship between the two, however, is likely to be complex, context-specific, and driven by underlying correlates [e.g., @blader_whats_2014; @cheng_dominance_2019; @kantner_identifying_2010; @von_rueden_roots_2014].

<!-- MALE BIAS -->
Substantial evidence suggests a near universal male-bias in community leadership across cultures [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse nonhuman social species [@smith_obstacles_2020; @tiger_female_1978]. Sex, however, is confounded with factors such as social network size, economic specialization, and education, that are also associated with leadership. After controlling for these factors, some studies have found that sex is not a strong predictor of community leadership [@yanca_female_2004; @garfield_investigating_2020; @von_rueden_sex_2018].

<!-- ASCRIBED THINGS -->
In the context of institutionalized social stratification, such as among chiefdoms, community leadership is often highly influenced by heredity and formal class structures [@earle_how_1997; @stanish_evolution_2004; @redmond_introduction_1998]. Managerial elites maintain influence over community decisions and are endowed with a high degree of social status [@stanish_evolution_2010]. Within state-level societies, including nonindustrial kingdoms, leadership positions are proscribed and institutionalized. Ruling classes, parties, or political bodies monopolize political influence and maintain control through a variety of mechanisms including military force, ideological systems, and control of information [@cohen_comparative_1967; @bodley_cultural_2011].

@@

## The benefits and costs of leadership across cultures and contexts
<!-- COSTS: OPPURTUNITY COSTS -->
Evolutionary models generally assume that, on average, the fitness benefits of leadership outweigh the fitness costs. The costs, however, have received less attention. Fulfilling group responsibilities and mediating conflicts between individuals can be physically and socially costly, and some decisions will upset some followers [@wiessner_power_2010]. In the absence of institutionalized authority, all group members are equipped to challenge leaders directly or indirectly using a variety of leveling mechanisms including physical aggression, gossip, and ridicule [@boehm_purposive_2008; @hess_gossip_2017]. Physically formidable individuals with strong social networks are likely to be able to bear the costs of leadership more so than their counterparts [@glowacki_leadership_2015].

@@

<!-- BENEFITS: 'MODEL' OF POTENTIAL BENEFIT -->
Although, empirically, leadership is *associated* with benefits, evolutionary models require that leaders cause or generate benefits. Potential mechanisms include that leaders (1) facilitate collective actions that yield net benefits to themselves or their close kin; (2) claim a fee for their services, such as a tax or greater share of returns; (3) receive reciprocal exchange in other currencies from followers; or (4) gain other social or reproductive benefits by signaling their high qualities [@von_rueden_leadership_2015]. Leaders can also promote policies that either align with their individual interests or that are not especially costly for themselves, their kin, or their social partners [@garfield_evolutionary_2019; @kantner_identifying_2010; @hagen_leadership_2019], even if 'just for a while.'

@@

We seek to provide a systematic and near comprehensive view of the qualities and functions of leaders, and the costs and benefits of leadership and followership, from a representative cross-cultural sample of nonindustrial populations. We then explore how these dimensions of leadership vary within populations across social contexts and across populations with different subsistence strategies, as well as by leader sex and continental region. We aim to identify *features* of the qualities and functions of leaders (e.g., do particular qualities or functions tend to covary or cluster together?). Such features might lend support to existing theories; for instance they might correspond to qualities associated with "prestige" or "dominance" [@henrich_evolution_2001] or functions associated with conflict or cooperation [@gavrilets_collective_2015; @glowacki_evolutionary_2017; @hooper_theory_2010]. Finally, we hope to synthesize the empirically-derived features of leadership with evolutionary theories of social organization to propose essential elements of leadership.

@@

We use the database of `r nrow(all_data)` ethnographic paragraphs (termed text records) on leadership, described in @garfield_evolutionary_2019 (https://doi.org/10.5281/zenodo.2541999), to code entirely new variables on the functions and qualities of leaders and the costs and benefits of leadership and followership. The text records were extracted from the 60-culture Probability Sample Files (PSF) of the electronic Human Relations Area Files (eHRAF) [@naroll_proposed_1967]. The PSF, which comprises over `r trunc(sum(leader_cult['ehraf_pages'])/1000)`K pages of digitized text, aims to mitigate Galton's problem of cultural non-independence by randomly sampling of one culture with high quality ethnographic coverage from each of 60 culture regions.

@@

Our first goal was to assess if discovery of evidence for each of our leadership dimensions was biased by date of publication, total pages of ethnography for a culture, and sex of the ethnographer(s). Our second goal was to assess the proportion of text records and cultures that provided evidence for each of our coded dimensions. We computed the proportion of text records supporting each dimension, and their 95% confidence intervals (CIs), using intercept-only logistic mixed effects models with random intercepts for document authors nested within cultures, fit with the lme4 package [@bates2015]. We then computed the proportion of cultures with at least one text record with evidence for each dimension, with 95% CIs computed using a cluster bootstrap.

@@

The geographic distribution of the cultures in our sample is displayed in Figure \@ref(fig:geographic) (see also Figure S\@ref(fig:subsisttree)). The database of text record comprised `r nrow(all_data)` text records from `r length(unique(leadershipdata::documents$d_ID))` unique documents describing `r length(unique(all_data$d_culture))` cultures.

@@

Subsistence strategy, group structure and continental region were confounded. Hunter-gatherer text records often referred to residential and kin leadership, but rarely to supracommunity leadership, whereas agricultural text records rarely referred to residential leadership but often referred to supracommunity leadership. Community level political groups were frequent across all subsistence types. African text records were predominantly from horticultural societies whereas North American text records were predominantly from hunter-gatherer societies. See Figures \@ref(fig:grouptype) and S\@ref(fig:regionsubsistence), and Table S\@ref(tab:fourfactorsumm).

@@

We investigated if there were biases in the discovery of evidence for any of our `r length(all_study_vars)` leader dimensions by publication date of source ethnographic texts, female gender of any author or co-author, or total pages of ethnography from each culture. Because we found minimal evidence of bias, we do not control for any bias variables in subsequent analyses. For details, see the *Bias* section of the SI.

@@

For each of the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded dimensions (Tables S\@ref(tab:qualfuncvars) and S\@ref(tab:costbenefitvars)), we computed the proportion of text records (95% CI) that provided support and then computed the proportion of cultures with at least one text record providing support for that dimension (as described in the Methods section). See Figure \@ref(fig:varsupport). For a table of all coefficients and 95% CIs, see Table S\@ref(tab:glmmcoefs).

@@

The most common functions of leaders, documented in over 70% of cultures, were resolving conflicts, providing miscellaneous social functions, organizing cooperative activities, and providing counsel or direction to followers. In about half of cultures sampled here leaders punished group members and managed economic systems. Functions notable for their rarity included controlling group movement and immigration.

The most common qualities of leaders were having high social status (over 90% of cultures), being experienced or accomplished, and being described as knowledgeable or intelligent (approximately 80% of cultures). Qualities notable for their relative infrequency included physical formidability, having many social contacts, and being feared (documented in about 35% of cultures) and charisma and fairness (documented in about 20% of cultures).

@@

Evidence for leader and follower costs and benefits were estimated similarly to the leader qualities and functions described above. Common leader benefits included greater access to material resources (documented in `r leader_benefit_material`% of cultures) and increased social status (documented in `r leader_benefit_status`% of cultures). Evidence for mating benefits and social services benefits was also relatively common. In general, the costs tracked the benefits. Leaders were often described as losing their high social status (documented in `r leader_cost_status`% of cultures) and greater access to resources (documented in `r leader_cost_material`% of cultures). See Figure \@ref(fig:varsupportcb).

Followers on the other hand most commonly benefited from leadership systems by receiving material resources (documented in `r follower_benefit_material`% of cultures) but otherwise by receiving social services (documented in `r follower_benefit_social`% of cultures). Evidence of mating and fitness benefits was relatively rare for followers. Followers were most frequently described as incurring costs in the form of material resources lost (documented in `r follower_cost_material`% of cultures) and in providing social services (documented in `r follower_cost_social`% of cultures). See Figure \@ref(fig:varsupportcb).

@@

We first examined the leadership dimensions with the most evidence, i.e., the  `r commondim_n` for which evidence was found in at least 60% of cultures. Of these, `r universals_n` dimensinos did not meaningfully vary by continental region, subsistence strategy, group context, or leader sex, and are therefore candidates for universal dimensions of human leadership. See Table \@ref(tab:leaderuniversals).

@@

We then examined the `r vary_n` (out of `r nrow(multi_allvars)` total) leadership dimensions that exhibited important variation by one or more of our four factors, i.e., whose AIC values indicated improved fits over the intercept-only models (Tables S\@ref(tab:variablevars) and S\@ref(tab:universalvars)). To determine which of the four factors contributed to improved fit, we dropped each factor in turn and computed the change in AIC. For most of these `r vary_n` varying dimensions of leadership, one factor had a noticeably bigger effect on fit than the others. Variation in group context, in particular, was associated with variation in almost half of the dimensions (Figure \@ref(fig:heatmapdrop1)A).

@@

For each of our four factors, we then investigated which levels of the factor was associated with higher or lower levels of evidence for the leadership dimension. For instance, there were three leadership dimensions in which subsistence strategy played an important role: *No coercive authority*, *Provide subsistence*, and *Distribute resources* (See Figure \@ref(fig:heatmapdrop1)A). Hunter-gatherers had particularly high levels of evidence for *No coercive authority*, relative to the other subsistence strategies (and averaging over the levels of the other three factors), and Pastoralists had relatively low evidence for *Distribute resources* (see Figure S\@ref(fig:heatmapsubsis)). For Group context, see Figure S\@ref(fig:heatmapgroups); for Region, see Figure, S\@ref(fig:heatmapregion); and for Leader sex, see Figure S\@ref(fig:heatmapsex).

@@

*High status* was the most frequently identified leader quality in the sample, with `r statusPercTxt`% of text records and `r statusPercCult`% of cultures containing evidence that leaders had high social status. To explore factors that distinguished these records from those with no evidence for high status, we first created a document-term matrix (DTM) of all "informative" words in our corpus of texts and the frequency with which they occurred in each text record (see SI for details). We then fit an elasticnet logistic regression model (with the lasso penalty, $\alpha = 1$) of *High status* (0/1) as a function of the frequencies of all `r ncol(leader_dtm)-1` words. Words that were strong positive predictors epitomized the semantic content of the text records which provided evidence of high status leaders. These included *`r paste(rev(highstatus_plot$data$Word)[1:5], collapse=', ')`* and *`r rev(highstatus_plot$data$Word)[6]`*, which were the strongest predictive terms. Notable weak-predictor terms included, *influence*, *authority*, *headman*, and *chief*, and *economic* was a negative predictor. See Figure \@ref(fig:highstatusplot)A.

@@

*High status* meaningfully differed among the different levels of group context, subsistence strategies, continental regions, and leader sex, with the most important factors being region followed by group context (see Figure \@ref(fig:heatmapdrop1)A). Examining the estimated marginal means, South America and East Eurasia had relatively high levels of evidence compared to Africa and North America, and kin and religious groups had relatively high levels of evidence compared to political groups. Nevertheless, the differences were quite modest. See Figure S\@ref(fig:groupstatus).

@@

Using all other quality and function dimensions as predictors identified *Punishment* as the strongest predictor of evidence for *Coercive authority*. *No coercive authority* was mostly strongly predicted by *Humility*. See Figures \@ref(fig:elasticcoercive)B,D.

@@

There were modest or no correlation between features, after aggregating leadership dimensions (see Figure S\@ref(fig:featurecor)). We then fit separate logistic mixed effects models of each of the `r nrow(feature_models)` features as a function of group context, subsistence strategy, continental region, and leader sex. See the SI for details.

@@

This exploratory study systematically evaluated evidence for `r length(all_study_vars)` dimensions of leadership in a stratified random sample of the ethnographic record of `r nrow(leader_cult)` largely nonindustrial cultures that included hunter-gatherers, pastoralists, horticulturalists, and agriculturalists. Results revealed universal dimensions of human leadership as well as important variation by group context, subsistence type, continental region, and sex of the leader.

@@

We found strong evidence that several dimensions of leadership were universal across cultures, and across group contexts within cultures. Leaders are seen as high status, knowledgeable or intelligent, and experienced or accomplished in about 80% or more of cultures, and function to resolve conflict, organize cooperation, and provide counsel or direction in over 70% of cultures (see Figure \@ref(fig:varsupport)). Leaders benefited materially, reproductively, or socially in 50% or more of cultures (see Figure \@ref(fig:varsupportcb)).

@@

These potentially universal dimensions of leadership correspond to those highlighted by several evolutionary theories that have drawn heavily on select ethnographic cases [@garfield_review_2019]. They partially correspond to the prestige model [@henrich_big_2015; @henrich_evolution_2001], which emphasizes knowledge, skill and generosity, but they do not include a key feature of that model: emulation of prestigious leaders. We labeled one of our leader quality features "cultural conformity", however, in which leaders are culturally conservative and exemplify desirable traits such as honesty, humility, and "proper behavior" (Figure \@ref(fig:cluster)). This feature might correspond, at least in part, to leaders who serve as models for their followers [see also @garfield_investigating_2020].

The universal dimensions also partially correspond to the service-for-prestige model [@price_evolution_2014; @price_service-for-prestige_2015] in which leaders receive special status in exchange for the services they provide followers, but that model does not emphasize the concrete material benefits seen here. The absence of *High status* and *Aggressiveness* on this list of universals corresponds to the reverse dominance hierarchy model [@boehm_egalitarian_1993], which emphasizes knowledge, intelligence, conflict resolution, generosity, and status leveling mechanisms among largely egalitarian societies.

@@

Evidence for several theoretically important leadership dimensions was relatively rare in the ethnographic record. The nonhuman animal leadership literature, for example, commonly focuses on group movement [@couzin_effective_2005], and "movement" was one of four leadership dimensions that @smith_leadership_2016 identified as comparable among small-scale human societies and nonhuman animal societies, yet movement or migration was identified as a leader function in only `r movement_pct`% of cultures, of which the vast majority (`r movement_hg_pct`%) were hunter-gatherers.

>>>>

Anthropologists and sociologists realized early on that leadership and followership were critical to understanding human psychology, social organization, and culture [e.g., @mumford_origins_1906; @firth1927economic; @morgan_ancient_1877; @myres_influence_1917]. Leaders are documented among every ethnographically observed society [@brown_human_1991; @lewis_leaders_1974], and in diverse contexts, such as kin groups [@dussart2000politics], ritual [@singh_cultural_2017], economic groups [@macfarlan_competitive_2012], conflicts between groups [@glowacki_evolutionary_2017], and nonindustrial political groups [@cohen_comparative_1967]. For review, see @garfield_review_2019.

Despite decades of research and theorizing on leadership, to our knowledge there is no systematic investigation of leader functions and qualities, and the costs and benefits of leadership and followership, across a representative sample of nonindustrial societies. Moreover, outside of anthropology, most scholarship on leadership is based on data from Western or Westernized, postindustrial societies [@von_rueden_leadership_2015].

<!-- WEIRD PEOPLE PROBLEMS -->
There is now a broad consensus on the importance of data from the full range of human cultural diversity. Since the seminal WEIRD people paper [@henrich_weirdest_2010], many social scientists have endevored to obtain data from "non-WEIRD" populations. Dichotomizing populations and their psychologies and social dynamics as WEIRD or non-WEIRD, however, is a mistake.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] As anthropology has conclusively demonstrated, there is enormous diversity across societies that does not remotely resemble a dichotomy.

@@

## The functions of leadership across cultures and contexts
<!-- INTRO -->
@roscoe_social_2009 proposed leadership in small-scale societies is structured by the joint demands of facilitating within-group conflict resolution and cooperation, and signaling coalitional strength to rival groups. Among many small-scale societies, which generally lack institutionalized political structures, leadership at the community level is typically informal and leaders emerge in response to group demands [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. Theoretical literature suggests these demands most commonly arise in the contexts of collective actions, resolving conflicts between families, and punishing norm violations [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2020]. Leaders also play important roles in ritual contexts, including initiations [@singh_cultural_2017], facilitating marriages [@walker_evolutionary_2011], and organizing feasts [@wiessner_food_1996]. Between-group interactions, including conflict and cooperation, are also common among nonindustrial societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @apicella_social_2012].

@@

<!-- MANAGEMENT FUNCTION IN MORE CULTUARLL COMPLEX -->
Stronger authority of community leaders, leadership beyond the community level, and institutionalized managerial roles are associated with sedentarization, defensibility of resources, and social stratification [@johnson_evolution_1987; @johnson_organizational_1982; @kaplan_evolutionary_2009; @ames_hierarchies_1985]. Among horticulturalists and agriculturalists, managerial leadership can promote more efficient use of shared resources such as water reserves and cooperative labor [e.g., @von_rueden_leadership_2014]. Pastoralists and coastal populations often face similar pressures concerning grazing lands and fishing access [e.g., @widmer_evolution_1988; @stevens_sherpa_1990]. 
<!-- In the wake of economic intensification, the functions of leadership becomes increasingly associated with managing economic systems and distribution of resources.  -->
Defensibility of resources creates increased opportunities for resource management and is often associated with increased territoriality and inter-group conflicts [@glowacki_evolutionary_2017]. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military operations [@lopez_evolutionary_2017].

@@

<!-- STATUS VS LEADERSHIP -->
Leaders have disproportionate influence, whereas status is often conceptualized as a relative indicator of the social value of the individual by the group and of their relative access to contested resources [e.g., @blader_whats_2014; @von_rueden_roots_2014]. Leaders are generally expected to be high status [e.g., @von_rueden_leadership_2015; @van_vugt_genes_2020; @cheng_dominance_2019; @buss_human_2020]. The causal relationship between the two, however, is likely to be context-specific and driven by underlying correlates [e.g., @blader_whats_2014; @cheng_dominance_2019; @von_rueden_roots_2014].

<!-- MALE BIAS -->
Substantial evidence suggests a near universal male-bias in community leadership [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse nonhuman social species [@smith_obstacles_2020; @tiger_female_1978]. Sex, however, is confounded with factors such as social network size, economic specialization, and education, that are also associated with leadership. After controlling for these factors, some studies have found that sex is not a strong predictor of community leadership [@yanca_female_2004; @garfield_investigating_2020; @von_rueden_sex_2018].

<!-- ASCRIBED THINGS -->
In the context of institutionalized social stratification, such as among chiefdoms, community leadership is often highly influenced by heredity and class structures [@earle_how_1997; @stanish_evolution_2004; @redmond_introduction_1998]. Managerial elites maintain influence over community decisions and are endowed with a high degree of social status [@stanish_evolution_2010]. Within state-level societies, including nonindustrial kingdoms, leadership positions are proscribed and institutionalized. Ruling classes, parties, or political bodies monopolize political influence and maintain control through a variety of mechanisms including military force, ideologies, and control of information [@cohen_comparative_1967; @bodley_cultural_2011].

@@

## The benefits and costs of leadership across cultures and contexts
<!-- COSTS: OPPURTUNITY COSTS -->
Evolutionary models generally assume that, on average, the fitness benefits of leadership outweigh the fitness costs. The costs, however, have received less attention. Fulfilling group responsibilities and mediating conflicts can be physically and socially costly, and some decisions will upset some followers [@wiessner_power_2010]. In the absence of institutionalized authority, all group members are equipped to challenge leaders directly or indirectly using a variety of leveling mechanisms including physical aggression, gossip, and ridicule [@boehm_purposive_2008; @hess_gossip_2017]. Physically formidable individuals with strong social networks are likely to be able to bear the costs of leadership more so than their counterparts [@glowacki_leadership_2015].

@@

<!-- BENEFITS: 'MODEL' OF POTENTIAL BENEFIT -->
Although, empirically, leadership is *associated* with benefits, evolutionary models require that leaders cause or generate benefits. Potential mechanisms include that leaders (1) facilitate collective actions that yield net benefits to themselves or close kin; (2) claim a fee or greater share of returns for their services; (3) receive reciprocal exchange in other currencies; or (4) gain other social or reproductive benefits by signaling their high qualities [@von_rueden_leadership_2015]. Leaders can also promote policies that either align with their individual interests or that are not especially costly for themselves, their kin, or their social partners [@garfield_evolutionary_2019; @kantner_identifying_2010; @hagen_leadership_2019], even if 'just for a while.'

@@

We seek to provide a systematic and near comprehensive view of the qualities and functions of leaders, and the costs and benefits of leadership and followership, from a representative sample of nonindustrial populations. We then explore how these dimensions of leadership vary within populations across social contexts and across populations with different subsistence strategies, as well as by leader sex and continental region. We aim to identify *features* of the qualities and functions of leaders (e.g., do particular qualities or functions tend to covary or cluster together?). Such features might lend support to existing theories; for instance they might correspond to qualities associated with "prestige" or "dominance" [@henrich_evolution_2001] or functions associated with conflict or cooperation [@gavrilets_collective_2015; @glowacki_evolutionary_2017; @hooper_theory_2010]. Finally, we hope to synthesize empirically-derived features of leadership with evolutionary theories of social organization to propose essential elements of leadership.

@@

We use the database of `r nrow(all_data)` ethnographic paragraphs (termed text records) on leadership, described in @garfield_evolutionary_2019 (https://doi.org/10.5281/zenodo.2541999), to code entirely new variables on the functions and qualities of leaders and the costs and benefits of leadership and followership. The text records were extracted from the 60-culture Probability Sample Files (PSF) of the electronic Human Relations Area Files (eHRAF) [@naroll_proposed_1967]. The PSF, which comprises over `r trunc(sum(leader_cult['ehraf_pages'])/1000)`K pages of digitized text, aims to mitigate Galton's problem of cultural non-independence by randomly sampling of one culture with high quality ethnographic coverage from each of 60 regions.

@@

Our first goal was to assess if discovery of evidence for each leadership dimensions was biased by date of publication, total pages of ethnography for a culture, and sex of the ethnographer(s). Our second goal was to assess the proportion of text records and cultures that provided evidence for each of our coded dimensions. We computed the proportion of text records supporting each dimension, and their 95% confidence intervals (CIs), using intercept-only logistic mixed effects models with random intercepts for document authors nested within cultures, fit with the lme4 package [@bates2015]. We then computed the proportion of cultures with at least one text record with evidence for each dimension, with 95% CIs computed using a cluster bootstrap.

@@

The geographic distribution of the cultures in our sample is displayed in Figure \@ref(fig:geographic) (see also Figure S\@ref(fig:subsisttree)). The database comprised `r nrow(all_data)` text records from `r length(unique(leadershipdata::documents$d_ID))` documents describing `r length(unique(all_data$d_culture))` cultures.

@@

Subsistence strategy, group structure and continental region were confounded. Hunter-gatherer text records often referred to residential and kin leadership, but rarely to supracommunity leadership, whereas agriculturalist text records rarely referred to residential leadership but often referred to supracommunity leadership. Community level political groups were frequent across all subsistence types. African text records were predominantly from horticultural societies whereas North American text records were predominantly from hunter-gatherer societies. See Figures \@ref(fig:grouptype) and S\@ref(fig:regionsubsistence), and Table S\@ref(tab:fourfactorsumm).

@@

We investigated if there were biases in the discovery of evidence for any of our `r length(all_study_vars)` leader dimensions by document publication date, female gender of any author or co-author, or total pages of ethnography from each culture. Because we found minimal evidence of bias, we do not control for any bias variables in subsequent analyses. For details, see the *Bias* section of the SI.

@@

For each of the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded dimensions (Tables S\@ref(tab:qualfuncvars) and S\@ref(tab:costbenefitvars)), we computed the proportion of text records (95% CI) that provided support and then computed the proportion of cultures with at least one text record providing support for that dimension (see Methods section). See Figure \@ref(fig:varsupport) and Table S\@ref(tab:glmmcoefs).

@@

The most common leader functions, documented in over 70% of cultures, were resolving conflicts, providing social functions, organizing cooperation, and providing counsel or direction to followers. In about half of cultures sampled here leaders punished group members and managed economic systems. Controlling group movement and immigration were notably rare functions.

The most common leader qualities were having high social status (over 90% of cultures), being experienced or accomplished, and being knowledgeable or intelligent (approximately 80% of cultures). Notably rare qualities included physical formidability, having many social contacts, and being feared (documented in about 35% of cultures) and charisma and fairness (documented in about 20% of cultures).

@@

Evidence for leader and follower costs and benefits were estimated similarly to the leader qualities and functions described above. Common leader benefits included greater access to material resources and increased social status (documented in `r leader_benefit_material`% and `r leader_benefit_status`% of cultures, respectively). Evidence for mating benefits and social services benefits was also relatively common. In general, the costs tracked the benefits. Leaders were often described as losing their high social status and greater access to resources (documented in `r leader_cost_status`% and `r leader_cost_material`% of cultures, respectively). See Figure \@ref(fig:varsupportcb).

Followers on the other hand most commonly benefited from leadership systems by receiving material resources (documented in `r follower_benefit_material`% of cultures) but otherwise by receiving social services (documented in `r follower_benefit_social`% of cultures). Evidence of mating and fitness benefits was relatively rare for followers. Followers were most frequently described as incurring costs in the form of material resources lost and in providing social services (documented in `r follower_cost_material`% and `r follower_cost_social`% of cultures, respectively). See Figure \@ref(fig:varsupportcb).

@@

We first examined the leadership dimensions with the most evidence, i.e., the  `r commondim_n` for which evidence was found in at least 60% of cultures. Of these, `r universals_n` dimensions did not meaningfully vary by continental region, subsistence strategy, group context, or leader sex, and are therefore candidates for universal dimensions of human leadership. See Table \@ref(tab:leaderuniversals).

@@

We then examined the `r vary_n` (out of `r nrow(multi_allvars)` total) leadership dimensions that exhibited important variation by one or more of our four factors, i.e., whose AIC values indicated improved fits over the intercept-only models (Tables S\@ref(tab:variablevars) and S\@ref(tab:universalvars)). To determine which of the four factors contributed to improved fit, we dropped each factor in turn and computed the change in AIC. For most of these `r vary_n` varying dimensions of leadership, one factor had a noticeably bigger effect on fit than the others. Variation in group context, in particular, was associated with variation in almost half of the dimensions (Figure \@ref(fig:heatmapdrop1)).

@@

For each of our four factors, we then investigated which levels of the factor was associated with higher or lower levels of evidence for the leadership dimension. For instance, there were three leadership dimensions in which subsistence strategy played an important role: *No coercive authority*, *Provide subsistence*, and *Distribute resources* (See Figure \@ref(fig:heatmapdrop1)). Hunter-gatherers had particularly high levels of evidence for *No coercive authority*, relative to the other subsistence strategies (and averaging over the levels of the other three factors), and Pastoralists had relatively low evidence for *Distribute resources* (see Figure S\@ref(fig:heatmapsubsis)). For Group context, see Figure S\@ref(fig:heatmapgroups); for Region, see Figure, S\@ref(fig:heatmapregion); and for Leader sex, see Figure S\@ref(fig:heatmapsex).

@@

*High status* was the most frequently identified leader quality in the sample (`r statusPercTxt`% of text records and `r statusPercCult`% of cultures provided evidence of high status leaders). To explore factors that distinguished these records from those with no evidence for high status, we first created a document-term matrix (DTM) of all "informative" words in our corpus of texts and the frequency with which they occurred in each text record (see SI for details). We then fit an elasticnet logistic regression model (with the lasso penalty, $\alpha = 1$) of *High status* (0/1) as a function of the frequencies of all `r ncol(leader_dtm)-1` words. Words that were strong positive predictors epitomized the semantic content of the text records which provided evidence of high status leaders. These included *`r paste(rev(highstatus_plot$data$Word)[1:5], collapse=', ')`* and *`r rev(highstatus_plot$data$Word)[6]`*, which were the strongest predictive terms. Notable weak-predictor terms included, *influence*, *authority*, *headman*, and *chief*, and *economic* was a negative predictor. See Figure \@ref(fig:highstatusplot)A.

@@

*High status* meaningfully differed by levels of group context, subsistence strategy, continental region, and leader sex, with the most important factors being region followed by group context (see Figure \@ref(fig:heatmapdrop1)). Examining the estimated marginal means, South America and East Eurasia had relatively high levels of evidence compared to Africa and North America, and kin and religious groups had relatively high levels of evidence compared to political groups. Nevertheless, the differences were quite modest. See Figure S\@ref(fig:groupstatus).

@@

Using all other quality and function dimensions as predictors identified *Punishment* as the strongest predictor of evidence for *Coercive authority*. *No coercive authority* was most strongly predicted by *Humility*. See Figures \@ref(fig:elasticcoercive)B,D.

@@

There were modest or no correlation between features, after aggregating leadership dimensions (see Figure S\@ref(fig:featurecor)). We fit separate logistic mixed effects models of each of the `r nrow(feature_models)` features as a function of group context, subsistence strategy, continental region, and leader sex, reported in the SI.

@@

This exploratory study systematically evaluated evidence for `r length(all_study_vars)` dimensions of leadership in a stratified random sample of the ethnographic record of `r nrow(leader_cult)` largely nonindustrial cultures including hunter-gatherers, pastoralists, horticulturalists, and agriculturalists. Results revealed universal dimensions of leadership as well as important variation by group context, subsistence type, continental region, and leader sex.

@@

We found strong evidence that several dimensions of leadership were universal across cultures and across contexts within cultures. Leaders are seen as high status, knowledgeable or intelligent, and experienced or accomplished in about 80% or more of cultures, and function to resolve conflict, organize cooperation, and provide counsel or direction in over 70% of cultures (see Figure \@ref(fig:varsupport)). Leaders benefited materially, reproductively, or socially in 50% or more of cultures (see Figure \@ref(fig:varsupportcb)).

@@

These potentially universal dimensions of leadership correspond to those highlighted by several evolutionary theories that have drawn heavily on select ethnographic cases [@garfield_review_2019]. They partially correspond to the prestige model [@henrich_big_2015; @henrich_evolution_2001], which emphasizes knowledge, skill and generosity, but they do not include a key feature of that model: emulation of prestigious leaders. We labeled one of our leader quality features "cultural conformity", however, in which leaders are culturally conservative and exemplify desirable traits such as honesty, humility, and "proper behavior" (Figure \@ref(fig:cluster)). This feature corresponds, at least in part, to leaders who serve as models for their followers [see also @garfield_investigating_2020].

The universal dimensions also partially correspond to the service-for-prestige model [@price_evolution_2014; @price_service-for-prestige_2015] in which leaders receive special status in exchange for providing services, but that model does not emphasize the concrete material benefits seen here. The absence of *High status* and *Aggressiveness* on this list of universals corresponds to the reverse dominance hierarchy model [@boehm_egalitarian_1993], which emphasizes knowledge, intelligence, conflict resolution, generosity, and status leveling mechanisms among largely egalitarian societies.

@@

Evidence for several theoretically important leadership dimensions was relatively rare in the ethnographic record. The nonhuman animal leadership literature, for example, commonly focuses on group movement [@couzin_effective_2005], and "movement" was one of four leadership dimensions @smith_leadership_2016 compared between small-scale human societies and nonhuman animal societies, yet movement or migration was identified as a leader function in only `r movement_pct`% of cultures, of which the vast majority (`r movement_hg_pct`%) were hunter-gatherers.
